---
title: "Project 1, Bioinformatics Institute, Statistics and R"
author: "Alexandra Gorbonos"
date: "6.11.2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Description

This project is the part of course "Statistics and R" in BI. The idea of the this project is to practice standard R functions and calculating some stats using data from the Olympic Games for the last 120 years.

## Task 1, Uploading files

Data is divided into several parts, we need to bind them.

```{r}
read_files <- function(path)
{
  total_path <- paste(getwd(), path, sep='')
  file_list <- list.files(path=total_path)
  dataset <- data.frame()
  print(file_list)
  for (i in 1:length(file_list)){
    temp_data <- read.csv(paste(total_path, file_list[i], sep="")) 
    dataset <- rbind(dataset, temp_data)
  }
  dataset
}


dataset <- read_files('/data_folder/')
```

## Task 2, Check data
We can use the command below to explore data:
```{r}
str(dataset)
```

Such variables as 'Height', 'Weight', 'Age' contain NA. To deal with NA in 'Height' and 'Weight' I decided to replace it with mean value, this solution was chosen to avoid missing large amount of data. In 'Age' column just a little part of table was NA, so I dropped it.
```{r}
dataset_without_NA <- dataset
mean_h <- mean(dataset_without_NA$Height[is.na(dataset_without_NA$Height) == FALSE])
dataset_without_NA$Height[is.na(dataset_without_NA$Height)] <- mean_h

mean_w <- mean(dataset_without_NA$Weight[is.na(dataset_without_NA$Weight) == FALSE])
dataset_without_NA$Weight[is.na(dataset_without_NA$Weight)] <- mean_w

dataset_without_NA <- dataset_without_NA[is.na(dataset_without_NA$Age) == FALSE,]
```

## Task 3, Where I need to find the youngest sportsmen (Male/Female) in the Olimpic Games, 1992

To check data i used: 
```{r}
check_s <- unique(dataset_without_NA$Sex)
print(check_s)
```
On this step I found one more problem with data: one participant has "G" in Sex column, so I googled this person and realized that he is male.

```{r}
dataset_without_NA$Sex[dataset_without_NA$Sex == 'G'] <- 'M'
```

To find the youngest sportsmen:
```{r}
M_data <- dataset_without_NA[dataset_without_NA$Sex == 'M',]
M_data[M_data$Age == min(M_data$Age),]

F_data <- dataset_without_NA[dataset_without_NA$Sex == 'F',]
F_data[F_data$Age == min(F_data$Age),]
```

## Task 4, Count mean and standart deviation for males and females height

I used dataframes from task 4 and mean/sd functions. 
```{r}
mean(M_data$Height)
sd(M_data$Height)

mean(F_data$Height)
sd(F_data$Height)
```

## Task 5, Count mean and standart deviation for females who played tennis in 2000

```{r}
mean(F_data$Height[(F_data$Year == 2000) & (F_data$Sport == 'Tennis')])
sd(F_data$Height[(F_data$Year == 2000) & (F_data$Sport == 'Tennis')])
```

## Task 6, Find the player with the heaviest weight in 2006, define the sport he participated in

```{r}
weights_2006 <- M_data$Weight[(M_data$Year == 2006)]
weights_2006 <- weights_2006[is.na(weights_2006) == FALSE]

M_data[(M_data$Weight == max(weights_2006)) & (M_data$Year == 2006), ]
```

## Task 7, The number of golden medals won by females on the period from  1980 to 2010

```{r}
F_data_medal <- F_data[is.na(F_data$Medal) == FALSE,]
F_data_medal <- F_data_medal[F_data_medal$Medal == 'Gold',]
F_data_medal <- F_data_medal[(F_data_medal$Year >= 1980) & (F_data_medal$Year <= 2010),]
sum((F_data_medal$Year >= 1980) & (F_data_medal$Year <= 2010))
```

## Task 8, How many times did John Aalberg participat in Olympic games in different years
```{r}
data_years <- dataset_without_NA[dataset_without_NA$Name == 'John Aalberg',]
years <- unique(dataset_without_NA$Year[dataset_without_NA$Name == 'John Aalberg'])
for (i in 1:length(years)){
  print(c(years[i], sum(data_years$Year == years[i])))
}
```

## Task 9, Count the numper of participants in different age groups

Possible age groups: [15-25), [25-35), [35-45), [45-55].
```{r}
dataset_without_NA_2008 <-  dataset_without_NA[dataset_without_NA$Year == 2008,]

data_15_25 <- dataset_without_NA_2008[dataset_without_NA_2008$Age >= 15 & dataset_without_NA_2008$Age < 25,]
print(length(unique(data_15_25$Name)))

data_25_35 <- dataset_without_NA_2008[dataset_without_NA_2008$Age >= 25 & dataset_without_NA_2008$Age < 35,]
print(length(unique(data_25_35$Name)))

data_35_45 <- dataset_without_NA_2008[dataset_without_NA_2008$Age >= 35 & dataset_without_NA_2008$Age < 45,]
print(length(unique(data_35_45$Name)))

data_45_55 <- dataset_without_NA_2008[dataset_without_NA_2008$Age >= 45 & dataset_without_NA_2008$Age <= 55,]
print(length(unique(data_45_55$Name)))
```

## Task10, The differense in number of sports between 1994 and 2002

```{r}
dataset_without_NA_2002 <-  dataset_without_NA[dataset_without_NA$Year == 2002,]
dataset_without_NA_1994 <-  dataset_without_NA[dataset_without_NA$Year == 1994,]

difference <- length(unique(dataset_without_NA_2002$Sport)) - length(unique(dataset_without_NA_1994$Sport))
print(difference)
```

## Task11, Top 3 countries for each kind of medal (Gold/Silver/Bronze)

Separately for the winter and summer olympics

Summer:
```{r}
country_list <- c()
medals_amount <- c()
Summer <- dataset_without_NA[dataset_without_NA$Season == 'Summer',]
Countries <- unique(Summer$Team)
Summer <- Summer[!is.na(Summer$Medal), ]
for (country in Countries){
  
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum(Summer$Team == country & Summer$Medal == 'Gold'))
}
summer_gold <- data.frame(country_list, medals_amount)
summer_gold <- summer_gold[order(-medals_amount),]
head(summer_gold, 3)
```

```{r}
country_list <- c()
medals_amount <- c()
Summer <- dataset_without_NA[dataset_without_NA$Season == 'Summer',]
Countries <- unique(Summer$Team)
Summer <- Summer[!is.na(Summer$Medal), ]
for (country in Countries){
  
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum(Summer$Team == country & Summer$Medal == 'Silver'))
}
summer_silver <- data.frame(country_list, medals_amount)
summer_silver <- summer_silver[order(-medals_amount),]
head(summer_silver, 3)
```

```{r}
country_list <- c()
medals_amount <- c()
Summer <- dataset_without_NA[dataset_without_NA$Season == 'Summer',]
Countries <- unique(Summer$Team)
Summer <- Summer[!is.na(Summer$Medal), ]
for (country in Countries){
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum(Summer$Team == country & Summer$Medal == 'Bronze'))
}
summer_bronze <- data.frame(country_list, medals_amount)
summer_bronze <- summer_bronze[order(-medals_amount),]
head(summer_bronze, 3)
```

Winter:
```{r}
country_list <- c()
medals_amount <- c()
Winter <- dataset_without_NA[dataset_without_NA$Season == 'Winter',]
Countries <- unique(Winter$Team)
Winter <- Winter[!is.na(Winter$Medal), ]
for (country in Countries){
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum(Winter$Team == country & Winter$Medal == 'Gold'))
}
winter_gold <- data.frame(country_list, medals_amount)
winter_gold <- winter_gold[order(-medals_amount),]
head(winter_gold, 3)
```

```{r}
country_list <- c()
medals_amount <- c()
Winter <- dataset_without_NA[dataset_without_NA$Season == 'Winter',]
Countries <- unique(Winter$Team)
Winter <- Winter[!is.na(Winter$Medal), ]
for (country in Countries){
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum(Winter$Team == country & Winter$Medal == 'Silver'))
}
winter_silver <- data.frame(country_list, medals_amount)
winter_silver <- winter_silver[order(-medals_amount),]
head(winter_silver, 3)
```

```{r}
country_list <- c()
medals_amount <- c()
Countries <- unique(Winter$Team)
Winter <- Winter[!is.na(Winter$Medal), ]
for (country in Countries){
  country_list <- c(country_list, country)
  medals_amount <- c(medals_amount, sum((Winter$Team == country) & (Winter$Medal == 'Bronze')))
  print(c(country, sum((Winter$Team == country) & (Winter$Medal == 'Bronze'))))
}
winter_bronze <- data.frame(country_list, medals_amount)
winter_bronze <- winter_bronze[order(-medals_amount),]
head(winter_bronze, 3)
```

## Task 12, Height_z_scores

```{r}
dataset_without_NA$Height_z_scores <- scale(dataset_without_NA$Height)
dataset_without_NA
```

## Task13, Min/Max normalization
```{r}
min_max_function <- function(x, na.rm = TRUE){
  return((x- min(x)) / (max(x)-min(x)))
}
dataset_without_NA$Height_min_max_scaled <- min_max_function(dataset_without_NA$Height)
```

## Task 14, Comparing Height, Weight and Age of males and females in winter Olympiad Games

#### Height
Normality test:
```{r}
F_data_winter <- F_data[F_data$Season == 'Winter',]
M_data_winter <- M_data[M_data$Season == 'Winter',]
shapiro.test(F_data_winter$Height[0:5000])
shapiro.test(M_data_winter$Height[0:5000])
```

Boxplot:
```{r}
FM_Height_boxplot <- ggplot(Winter, aes(x = as.factor(Sex), y=Height, group=Sex)) +
  geom_boxplot() + 
  scale_x_discrete(name = "Sex") +
  labs(title = "Fig 1. Height boxplots for males and females.") +
  theme_bw()
FM_Height_boxplot
```

H0: Males and Females have equal height.
H1: Males and Females have different height.
```{r}
t.test(x=F_data_winter$Height, y=M_data_winter$Height, alternative = c("two.sided"))
```

One more t test:
H0: Males and Females have equal height.
H1: Males are highter than females.
```{r}
t.test(x=F_data_winter$Height, y=M_data_winter$Height, alternative = c("less"))
```


#### Weight

Normality test:
```{r}
shapiro.test(F_data_winter$Weight[0:5000])
shapiro.test(M_data_winter$Weight[0:5000])
```

Boxplot:
```{r}
FM_Weight_boxplot <- ggplot(Winter, aes(x = as.factor(Sex), y=Weight, group=Sex)) +
  geom_boxplot() + 
  scale_x_discrete(name = "Sex") +
  labs(title = "Fig 1. Weight boxplots for males and females.") +
  theme_bw()
FM_Weight_boxplot
```

H0: Males and Females have equal weight.
H1: Males and Females have different weight.
```{r}
t.test(x=F_data_winter$Weight, y=M_data_winter$Weight, alternative = c("two.sided"))
```

One more t test:
H0: Males and Females have equal weight.
H1: Males have bigger weight than females.
```{r}
t.test(x=F_data_winter$Weight, y=M_data_winter$Weight, alternative = c("less"))
```


#### Age

Normality test:
```{r}
shapiro.test(F_data_winter$Age[0:5000])
shapiro.test(M_data_winter$Age[0:5000])
```

Boxplot:
```{r}
FM_Age_boxplot <- ggplot(Winter, aes(x = as.factor(Sex), y=Age, group=Sex)) +
  geom_boxplot() + 
  scale_x_discrete(name = "Sex") +
  labs(title = "Fig 1. Age boxplots for males and females.") +
  theme_bw()
FM_Age_boxplot
```

H0: Males and Females have the same age.
H1: Males and Females have different ages.
```{r}
t.test(x=F_data_winter$Age, y=M_data_winter$Age, alternative = c("two.sided"))
```

One more t test:
H0: Males and Females have the same age.
H1: Males are older than females.
```{r}
t.test(x=F_data_winter$Age, y=M_data_winter$Age, alternative = c("less"))
```

